---
title: "Examples in Randomized Experiment"
output:
  html_document:
    theme: simplex
    code_folding: show
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: no
    highlight: pygments
    df_print: paged
    fig_caption: no
  pdf_document:
    toc_depth: '3'
fontsize: 18pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Generating Data Set
We first generate a set of $N$ subjects who will participate in the experiment. Let $D_{i}\in\{0,1\}$ be a random variable representing the treatment assignment for subject $i$. $D_i=1$ means the subject $i$ is assigned to the treatment group and $D_i=0$ means the subject $i$ is assigned to the control group. Let $(Y_i(0),Y_i(1))$ be the (potential) outcomes of the subject $i$, e.g., time spent on the website under control and treatment intervention. 

Let $Y_i(0)\sim ~N(\mu_0,\sigma_0^2)$ and $Y_i(1) = N(\mu_1,\sigma^2_1)$.
```{r}
set.seed(1) # initialize the seed of the random number generator
N = 20 # total number of subjects participating in the experiment
mu0 = 10
sigma0 = 2
mu1 = 20
sigma1 = 2.5
Y0 = rnorm(N,mu0,sigma0^2)
Y1 = rnorm(N,mu1,sigma1^2)
Data = data.frame(cbind(Y0,Y1)) # potential outcome of the sample
names(Data) = c('Y0','Y1') 
Data
```
Note that the population average treatment effect is give by $\tau_{fs}=\bar{Y}_i(1)-\bar{Y}_i(0)=\frac{1}{N}\sum_{i=1}^NY_i(1)-\frac{1}{N}\sum_{i=1}^NY_i(0)$, which equals `mean(Y1-Y0)=``r mean(Y1-Y0)`. This is the objective. We want to infer this number from this simulated experiment.

## Assignment Procedure

### Assignment is Independent of the Potential Outcomes
Assume we flip a fair coin to determine the assignment of each subject (e.g., heads to the treatment and the tails to the control). To simulate the a random treatment assignment with this bernoual trials:
```{r message=FALSE, warning=FALSE}
library('mc2d')  
# there are many ways to generate 
# bernoulli random numbers: 1 means treatment
Index = rbern(N, 0.5) 
Index
```
In this case, we have `r sum(Index)` subjects being assigned to the treatment group. Hence, the observed averages in the treatment and control group are
```{r}
YobsT_avg = mean(Data[which(Index==1),'Y1'])
YobsC_avg = mean(Data[which(Index==0),'Y0'])
```
We have $\bar{Y}^{obs}_T=$ `r YobsT_avg` and $\bar{Y}^{obs}_C=$ `r YobsC_avg`, which leads to $\hat{\tau}_{fs}=\bar{Y}^{obs}_T-\bar{Y}^{obs}_C=$ `r YobsT_avg-YobsC_avg`. It is close but not equal to the true population average treatment effect `r mean(Y1-Y0)`. 

But if we reset the clock and execute this randomization procedure over and over again, then we have
```{r}
Rep = 10
Diff = rep(0,Rep)
for (i in seq(1,Rep,1))
{
 Index = rbern(N, 0.5)
 Diff[i] = mean(Data[which(Index==1),'Y1'])-mean(Data[which(Index==0),'Y0'])
 print(Diff[i])
  }
```
Then we take the averages of these values of our estimator $\hat{\tau}_{fs}=\bar{Y}^{obs}_T-\bar{Y}^{obs}_C$ obtained from each realized randomized assignment, then we have $\frac{1}{R}\sum_{r=1}^R\hat{\tau}_{fs}^{(r)}=$ `r mean(Diff)`. The number is closer to the true population average treatment effect `r mean(Y1-Y0)`, but still not equal.

If you reset the clock for more times and execute the randomization procedure, then we have
```{r}
Rep = 10000
Diff = rep(0,Rep)
for (i in seq(1,Rep,1))
{
 Index = rbern(N, 0.5)
 Diff[i] = mean(Data[which(Index==1),'Y1'])-mean(Data[which(Index==0),'Y0'])
}
print(mean(Diff))
```
Then you can see that the expected value of $\frac{1}{R}\sum_{r=1}^R\hat{\tau}_{fs}^{(r)}=$ `r mean(Diff)` is very close to the true population average treatment effect `r mean(Y1-Y0)`, as governed by the law of large number (LLN). 

### Assignment is NOT Independent of the Potential Outcomes
So far the assignment is independent of the potential outcomes, i.e., it is solely determined by random a coin flip. Now we try another assignment procedure, which is not independent of the potential outcome. Specifically, we assign the subject $i$ to treatment 1 if and only if $Y_i(1)/Y_i(0)+\epsilon_i>10$, where $\epsilon_i\sim \mathcal{N}(0,1)$.
```{r}
Rep = 10000
Diff2 = rep(0,Rep)
Ratio = Y1/Y0
for (i in seq(1,Rep,1))
{
 Index = (Ratio + rnorm(N,0,1)>10)
 Diff[i] = mean(Data[which(Index==1),'Y1'])-mean(Data[which(Index==0),'Y0'])
}
print(mean(Diff))
```
You can see that even if we reset the clock `r Rep` times, still the expected value of $\frac{1}{R}\sum_{r=1}^R\hat{\tau}_{fs}^{(r)}$ is different from the true population average treatment effect `r mean(Y1-Y0)` substantially.